---
title: "R filtering of Lake Ljøgottjern dataset"
author: "Anneke ter Schure"
output:
  pdf_document: default
  word_document: default
---

This Rmarkdown file uses the results from OBITools processing as input and further filters the metabarcoding data.
The code provided here is a mash-up from different sources, including the ForBio course: DNA Metabarcoding: Data processing and interpretation.
Special thanks and credit goes to the teachers: 
Eric Coissac
Frédéric Boyer
Youri Lammers

This script is certainly not the most concise or optimized, but it is functional :)
It includes the following steps:

- Reformatting of the data into samples, motus and reads data frames
- Checking the data
- Removing sequences with a maximum abundance in the negative controls
- Removing sequences with < 100% best identity match (plants), or 98% best identity match (mammals)
- Removal of sequences with < 10 reads in a PCR repeat
- Removal of PCR repeats with < 100 reads in total
- Removal of sequences present in < 2 PCR repeats & sequences with mean read count > mean read count in blanks
- Saving the filtered data

```{r}
# prepare the R environment
library(tidyverse)
```

Setting the filters
```{r}
# Filter thresholds to be adjusted according to experiment
bestid_thr = 1 # for best identity match with reference database
thrCount_motu <- 10 # minimal number of reads to keep a MOTU
thrCount_pcr <- 100 # minimal number of reads to keep a PCR replicate 
min_replicates = 2 # minimal number of replicates needed to keep a sample
```

```{r}
# load the data
raw = read.delim("../data/p1_assigned_trnl_embl_arct_r005.tab")
```
In 'raw' we have the OTUs as rows and all of the information (quality, taxonomy, readcounts etc. in columns)
Let's prepare a transformed version of the OTU table and seperate into specific tables:
 - samples
 - motus
 - reads

```{r}
# motus should be a dataframe with the metadata for the motu ids
# so we need a subset of the 'raw' dataframe without the readcounts
motus = select(raw, -(starts_with('sample')))

# and the opposite subset for the reads table
reads = select(raw, starts_with('sample'))

# change NAs (reads found for one database but not the other) into zeros 
# NOTE: this is not necessary for the mammal data
reads[is.na(reads)] = 0

# Samples should just be a dataframe with complete samplenames as rownames and as a column "Samples"
# so let's extract the sample names from the reads table
names = as.character(colnames(reads))

samplenames_split = strsplit(names,"[.]")

# save the last part as the full sample name
samplenames = sapply(samplenames_split,
                   function(x) x[length(x)])

samples = data.frame(sample = samplenames)
rownames(samples) = samplenames
# make sure the names match for the samples and reads tables
colnames(reads) = samplenames 
```
```{r}
# check the dimensions of the resulting dataframes
dim(motus)
dim(samples)
dim(reads)
```

Extracting the sample names and updating the samples data frame

```{r}
# get sample names
sample_names = as.character(samples$sample)

# Split the names on the underscores
sample_names_split = strsplit(sample_names,"_")
#head(sample_names_split)

# save the last part as replicate
replicate = sapply(sample_names_split,
                   function(x) x[length(x)])

# save the first part as the sample id
sample_id = sapply(sample_names_split,
                   function(x) x[1])

sample_type = sapply(sample_names_split,
                   function(x) x[2])
# label everyting as sample, except the controls
sample_type[grep(pattern = "^s",sample_type)]="sample"
sample_type[grep(pattern = "^pnc",sample_type)]="pcrneg"
sample_type[grep(pattern = "^enc",sample_type)]="extneg"
sample_type[is.na(sample_type)]="leakage" # in case we found some unused tags"

# builds a data frame with the sample descriptions
sample_desc = data.frame(name      = sample_names,
                         sample_id = sample_id,
                         replicate = replicate,
                         type      = factor(sample_type),
                         nonReplicating = FALSE # needed later on for filtering
                         )
rownames(sample_desc) = sample_desc$name

# Substitute the old sample description by the new one
samples = sample_desc
sample_desc = NULL # cleanup the workspace
```

Update the motus dataframe by adding some information about the reference databases
Note: this is only relevant for the plant sequence data

```{r}
sequence_type = rep("Unknown",nrow(motus))

# we have two database and we want to prioritise the arctborbryo; so lets check both against the bestid_thr and let arctborbryo overrule the embl hits
sequence_type[motus$`best_identity.trnl_r142` >= bestid_thr] = "embl" 
sequence_type[motus$`best_identity.arctborbryo.gh` >= bestid_thr] = "arctborbryo" 

motus$sequence_type=factor(sequence_type)

# also save the corresponding best_identity and related values for:
# "family_name", "genus_name", "species_name", "scientific_name"

# save the highest best identity by comparison of both
bestid = pmax(motus$`best_identity.arctborbryo.gh`,motus$best_identity.trnl_r142,na.rm = TRUE)
family = rep(NA,nrow(motus))
genus = rep(NA,nrow(motus))
species = rep(NA,nrow(motus))
scientific = rep(NA,nrow(motus))

# make sure that they are character vectors
motus = transform(motus, family_name = as.character(family_name),
                  a_family_name = as.character(a_family_name),
                  genus_name = as.character(genus_name),
                  a_genus_name = as.character(a_genus_name),
                  species_name = as.character(species_name),
                  a_species_name = as.character(a_species_name),
                  scientific_name = as.character(scientific_name),
                  a_scientific_name = as.character(a_scientific_name)
                  )

family = ifelse(motus$sequence_type == "embl", motus$family_name, motus$a_family_name)
genus = ifelse(motus$sequence_type == "embl", motus$genus_name, motus$a_genus_name)
species = ifelse(motus$sequence_type == "embl", motus$species_name, motus$a_species_name)
scientific = ifelse(motus$sequence_type == "embl", motus$scientific_name, motus$a_scientific_name)

# add the info to the motus table
motus$best_id=bestid
motus$family_name=family
motus$genus_name=genus
motus$species_name=species
motus$scientific_name=scientific
```

Basic descriptions of the samples
```{r}
#The dataset is constituted of 
nrow(reads) #PCRs replicates
ncol(reads) #MOTUs 

# Counting the reads per PCR reaction
reads_per_pcr = rowSums(reads)

# Counting the MOTUs per PCR reaction
motus_per_pcr = rowSums(reads > 0)

# Counting the reads per sample
reads_per_sample = aggregate(reads_per_pcr, list(sample_id), FUN=sum)
reads_per_sample[order(reads_per_sample$x, decreasing = FALSE),]

# Counting the MOTUs per sample
motus_per_sample = aggregate(motus_per_pcr, list(sample_id), FUN=sum)
motus_per_sample[order(motus_per_sample$x, decreasing = FALSE),]
```
```{r}
# taking a look at the distribution
par(mfrow=c(1,2))
hist(log10(reads_per_pcr),
     breaks=50,
     main = "Reads per PCR",
     xlab="Reads count (log scale)")
hist(motus_per_pcr,breaks=50,
     main = "MOTUs per PCR",
     xlab="MOTU count")
```

```{r}
# check the number of reads against the number of motus
plot(reads_per_pcr,motus_per_pcr,
     xlab="Read counts",
     ylab="MOTU counts",
     cex=0.5,
     col = samples$type,
     log="xy")

legend("bottomright",
       legend = levels(samples$type),
       fill = 1:length(levels(samples$type)))
```
Check the obiclean status
```{r}
obiclean_statussus = select(motus, starts_with("obiclean_status")) 
internal = obiclean_statussus =='i'
always_internal = apply(internal,MARGIN = 1,FUN = all,na.rm = TRUE) # margin was 2, but I think this version is transformed; make sure it is the MOTUS! (if numbers are higher than total pcr replicates, you're ok!)
table(always_internal)
```
```{r}
plot(reads_per_motu,
     motus$best_id,
     cex=0.5,
     col = always_internal + 1,
     xlab= "read counts per MOTU",
     ylab= "best id of the MOTU with ref DB",
     log="xy")
legend("bottomright",
       legend = c('H or S','Internal'),
       fill=1:2,
       cex=0.6)
```

# Filtering

## Identify artefacts

```{r}
#get #reads & #OTUs / sample 
plot(rowSums(reads), rowSums(reads>0), 
     col=as.factor(samples[rownames(reads), 'type']),
     log='xy', pch=16, xlab='#Reads', ylab='#OTUs')

legend('topleft', 
       legend = levels(as.factor(samples[rownames(reads), 'type'])),
       col=1:nlevels(as.factor(samples[rownames(reads), 'type'])), 
       pch=16, cex=0.5)
```

### Reagent contaminants; MOTUs of maximum abundance in the controls
Identification of OTUs of maximum abundance in the extraction and pcr negative controls
```{r, results="hide", warning=FALSE}
stopifnot(all(rownames(samples)==rownames(reads)))

maxInExtractionCtrl <- apply(reads[samples$type=='extneg',], MARGIN=2, function(x) max(x,na.rm = T))
maxInPCRCtrl        <- apply(reads[samples$type=='pcrneg',], MARGIN=2, function(x) max(x,na.rm = T)) 
maxInSamples        <- apply(reads[samples$type=='sample',], MARGIN=2, function(x) max(x,na.rm = T)) 

df <- data.frame(maxInExtractionCtrl, maxInPCRCtrl, maxInSamples)

#determining the type of problematic OTU
motus$bias <- c('extneg','pcrneg',NA)[apply(df, MARGIN=1, FUN=which.max)]

#adding OTUs infos
infosCols <- c("best_id", "family_name", "genus_name","species_name", "scientific_name", "bias", "sequence_type")

df <- cbind(df, motus[,infosCols])

#keeping only problematic sequences for display

df <- df[!is.na(df$bias),]
df <- df[order(df$maxInExtractionCtrl+df$maxInPCRCtrl, decreasing=T),]
```
Visualizing the level of contaminants in the samples
```{r}
stopifnot(all(colnames(reads)==motus$id))

if (dim(df)[1] > 1) {
  plot(rowSums(reads[, !is.na(motus$bias)]), rowSums(reads[,!is.na(motus$bias)]>0), 
     col=as.factor(samples[rownames(reads), 'type']),
     log='xy', pch=16, xlab='#Reads', ylab='#OTUs')

  legend('topleft', 
       legend = levels(as.factor(samples[rownames(reads), 'type'])),
       col=1:nlevels(as.factor(samples[rownames(reads), 'type'])), 
       pch=16, cex=0.5)
} else {
  plot(sum(reads[, !is.na(motus$bias)]), sum(reads[,!is.na(motus$bias)]>0), 
     col=as.factor(samples[rownames(reads), 'type']),
     log='xy', pch=16, xlab='#Reads', ylab='#OTUs')

  legend('topleft', 
       legend = levels(as.factor(samples[rownames(reads), 'type'])),
       col=1:nlevels(as.factor(samples[rownames(reads), 'type'])), 
       pch=16, cex=0.5)
}
```
Removing the contaminants
```{r}
table(motus$bias)

correctedCounts2 <- reads

#correctedCounts2[,!is.na(motus$bias) | motus$degraded] <- 0  ## removing contaminants and degraded motus; 

correctedCounts2[,!is.na(motus$bias)] <- 0  ## removing contaminants

# check how many are removed; FALSE means they're empty; TRUE means that they're kept
table(colSums(correctedCounts2) > 0)


plot(rowSums(reads), rowSums(correctedCounts2), xlab='#Reads before cleaning for contaminants',
     ylab='#Reads after cleaning for contaminants', main='Effect of contaminants removal\non #Reads',
     col=as.factor(samples[rownames(reads), 'type']), pch=16)
abline(a=0,b=1)

legend('topleft', 
       legend = levels(as.factor(samples[rownames(reads), 'type'])),
       col=1:nlevels(as.factor(samples[rownames(reads), 'type'])), 
       pch=16, cex=0.5)

plot(rowSums(reads>0), rowSums(correctedCounts2>0), xlab='#OTUs before cleaning for contaminants',
     ylab='#OTUs after cleaning for contaminants', main='Effect of contaminants removal\non #OTUs',
     col=as.factor(samples[rownames(reads), 'type']), pch=16)
abline(a=0,b=1)

legend('topleft', 
       legend = levels(as.factor(samples[rownames(reads), 'type'])),
       col=1:nlevels(as.factor(samples[rownames(reads), 'type'])), 
       pch=16, cex=0.5)
```
```{r}
sum(colSums(correctedCounts2) > 0) # total number of motus left
sum(rowSums(correctedCounts2)) # total number of reads left
```
```
```{r}
# updating the tables
correctedCounts2 = correctedCounts2[rowSums(correctedCounts2) > 0, colSums(correctedCounts2) > 0] # filter out empty sequences and samples
motus = motus[colnames(correctedCounts2),]
samples = samples[rownames(correctedCounts2),]
dim(correctedCounts2)
```

### Removing MOTUs < 100% match (plants) or 98% match (mammals)
```{r}
# keeping track of things
table(colSums(correctedCounts2) > 0) # TRUE are those that are kept

# determining which are below the bestid_thr and removing those from the reads
motus$below_bestid_thr = motus$best_id<bestid_thr
correctedCounts2[,motus$below_bestid_thr] <- 0 

# now check again, the number for TRUE should be smaller
table(colSums(correctedCounts2) > 0)

sum(colSums(correctedCounts2) > 0) # total number of motus left
sum(rowSums(correctedCounts2)) # total number of reads left
```
```{r}
# updating the tables
correctedCounts2 = correctedCounts2[rowSums(correctedCounts2) > 0,colSums(correctedCounts2) > 0] # filter out empty sequences and samples
motus = motus[colnames(correctedCounts2),]
samples = samples[rownames(correctedCounts2),]

dim(correctedCounts2)
```

### Removing MOTUs < 10 reads in a PCR repeat

```{r}
# determining which are below a certain read countand removing those from the reads
correctedCounts2[correctedCounts2<thrCount_motu] = 0
# check how many
table(colSums(correctedCounts2) > 0)
```
```{r}
# save and check what is left
reads1 = correctedCounts2[rowSums(correctedCounts2) > 0, colSums(correctedCounts2) > 0] # filter out empty sequences and samples
motus1 = motus[colnames(reads1),]
samples1 = samples[rownames(reads1),]

# check number of replicates, motus and reads
dim(reads1)
sum(colSums(reads1)) # total number of reads left
```

### Removing PCR repeats with < 100 reads total

```{r}
# removing those with low amount of reads / empty ones
l <- lapply(levels(as.factor(samples[rownames(reads1), 'type'])), function(st) {
  rowSums(reads1[samples[rownames(reads1), 'type']==st,])
})
names(l) <- levels(as.factor(samples[rownames(reads1), 'type']))

boxplot(l, las=2, cex=0.25, outpch=NA, main='#reads before filtering')
stripchart(l, vertical = T, 
           method="jitter", pch=16, cex=0.4, add=T)
abline(h=thrCount_pcr, col='red')
```
```{r}
## labelling samples that are below a threshold readcount
samples1$empty_PCR <- FALSE

stopifnot(all(rownames(samples1)==rownames(reads1)))
samples1$empty_PCR[rowSums(reads1)<thrCount_pcr] <- TRUE

## filtering out those that are "empty" (both PCR repeats as sequences)
correctedCounts3 = reads1[!samples1$empty_PCR, colSums(reads1) > 0]

reads2 = correctedCounts3[rowSums(correctedCounts3) > 0,] ## not sure this step is needed, but doesn't hurt
motus2 = motus[colnames(reads2),]
samples2 = samples[rownames(reads2),]
```

```{r}
# how many replicates, motus and reads do we have left?
dim(reads2)

sum(colSums(reads2) > 0) # total number of motus left
sum(rowSums(reads2)) # total number of reads left
```

### Final filtering steps
We still need to: 
- check of frequency of occurrence of MOTUs in PCR repeats 
- compare between the mean number of reads in blanks and in samples 

Prepare a list of blanks and samples
```{r}
# get the list of blanks 
# and calculate the mean number of reads per MOTU for these
blanklist = unique(samples2$sample_id[samples2$type != "sample"])
if (length(blanklist) == 0){
  blanks = 0
  blanks_mean = NULL
} else {
  blanks = reads2[grep(paste(blanklist,collapse="|"),rownames(reads2)),] 
  blanks_mean = apply(blanks,2,mean)
}

# get the list of samples
samplelist = unique(samples2$sample_id[samples2$type == "sample"])
```

For each sample check if MOTUs:
- are present in at least 2 replicates
- average count is higher than in blanks
```{r}
# initiate new reads dataframe and samplenameslist
reads3 = reads2[FALSE,]
keptreps = character()

# for each sample check if MOTUs:
# are present in at least 2 replicates
# average count is higher than in blanks
for (s in samplelist) {
  # make a subset only for that sample and create a binary table for the same subset
  temp = reads2[grep(s,rownames(reads2)),]
  temp_bin = temp
  temp_bin[temp_bin > 0] <- 1
  # check if there is more than one replicate to start with
  if (!is.null(dim(temp))) {
      # when there is less than a certain number of occurrences 
      # of the MOTU in the subset, set the MOTU to 0
      temp[,colSums(temp_bin) < min_replicates] <- 0 
      
      
      # calculate the mean per MOTU 
      temp_mean = apply(temp,2,mean)
      # check against the means in the blanks
      if (!is.null(blanks_mean)) {
          temp[blanks_mean>temp_mean] <- 0
      }
      reads3 = rbind(reads3, temp)
      keptreps <- c(keptreps, rownames(temp))
      
  } 
  else if (min_replicates == 1) {
    if (!is.null(blanks_mean)) {
          temp[blanks_mean>temp] <- 0
      }
      reads3 = rbind(reads3, temp)
      keptreps <- c(keptreps, rownames(reads2)[grep(s,rownames(reads2))])
  }
}

rownames(reads3) = keptreps
dim(reads3)
reads3 = reads3[,colSums(reads3)!= 0]
```
```{r}
# Check the dimensions of the resulting dataframes
dim(reads3)
sum(rowSums(reads3)) # total number of reads that are left
motus3 = motus2[colnames(reads3),]
samples3 = samples2[rownames(reads3),]
dim(motus3)
```
### Save the filtered data
```{r}
write.csv(motus3,file = paste0("../results/v2_TRNL_r005_", bestid_thr, "_count", thrCount_motu,"_cleanrepeats_motus.csv"))
write.csv(reads3,file = paste0("../results/v2_TRNL_r005_", bestid_thr, "_count", thrCount_motu,"_cleanrepeats_reads.csv")) 
write.csv(samples3,file = paste0("../results/v2_TRNL_r005_", bestid_thr, "_count", thrCount_motu,"_cleanrepeats_samples.csv"))
```
